### 2022-11-12 draft/draft.Rmd
still writing

```{r}
import::from(magrittr, `%>%`, `%<>%`)
import::from(foreach, foreach, `%dopar%`)
library(ggplot2)
library(ggraph)
library(mclust)
source('~/dev/pabm-grdpg/functions.R')
source('~/dev/manifold-block-models/functions.R')

@@

\begin{algorithm}
  \label{algo:sbm-em}
  \setstretch{1.5}
  \DontPrintSemicolon
  \SetAlgoLined
  \KwData{Adjacency matrix $A$, number of communities $K$}
  \KwResult{Estimated community label probabilities $\{\pi_{ik}\}$ for which each $\pi_{ik} = P(z_i = k \mid A)$, estimated community edge probabilities $\{\hat{\theta}_{k \ell}\}_K$}
  \caption{Approximate EM algorithm for the SBM}
  Initialize $\{\pi_{ik}\}$, $\{\theta_{k \ell}\}$.\;
  \While{$\|\nabla \ell\| > \epsilon$} {
    \For {$i = 1, ..., n$} {
      \For {$k = 1, ..., K$} {
      E-step: $\pi_{ik} \propto \exp \bigg( \sum_{j \neq i} \sum_{\ell} \pi_{j \ell} (A_{ij} \log \hat{\theta}_{k \ell} + (1 - A_{ij}) \log (1 - \hat{\theta}_{k \ell})) \bigg)$.\;
        \For {$\ell = 1, ..., K$} {
          M-step: $\hat{\theta}_{k \ell} = \frac{\sum_{i < j} A_{ij} \pi_{ik} \pi_{j \ell}}{\sum_{i < j} \pi_{ik} \pi_{j \ell}}$.\;
        }
      }
    }
  }
\end{algorithm}

@@

\begin{example}
\label{ex:uk-faculty-sbm}
Recall the UK faculty friendship network from example~\ref{ex:uk-faculty}. 
\end{example}

@@

\begin{algorithm}
  \label{algo:dcbm-em}
  \setstretch{1.5}
  \DontPrintSemicolon
  \SetAlgoLined
  \KwData{Adjacency matrix $A$, number of communities $K$}
  \KwResult{Estimated community label probabilities $\{\pi_{ik}\}$ for which each $\pi_{ik} = P(z_i = k \mid A)$, estimated community edge probabilities $\{\hat{\theta}_{k \ell}\}_K$, estimated degree factors $\hat{\omega}$.}
  \caption{Approximate EM algorithm for the DCBM.}
  Initialize $\{\pi_{ik}\}$, $\{\theta_{k \ell}\}_K$, $\hat{\omega}$.\;
  \While{$\|\nabla \ell\| > \epsilon$} {
    \For {$i = 1, ..., n$} {
      \For {$k = 1, ..., K$} {
      E-step: $\pi_{ik} \propto \exp \bigg( \sum_{j \neq i} \sum_{\ell} \pi_{j \ell} (A_{ij} \log (\hat{\theta}_{k \ell} \hat{\omega}_i \hat{\omega}_j) + (1 - A_{ij}) \log (1 - \hat{\theta}_{k \ell} \hat{\omega}_i \hat{\omega}_j)) \bigg)$.\;
        \For {$\ell = 1, ..., K$} {
          M-step for $\theta_{k \ell}$: $\hat{\theta}_{k \ell} = \frac{\sum_{i < j} A_{ij} \pi_{ik} \pi_{j \ell}}{\sum_{i < j} \pi_{ik} \pi_{j \ell} \hat{\omega}_i \hat{\omega}_j}$.\;
        }
      }
      M-step for $\omega_i$: $\hat{\omega}_i = \frac{\sum_{j \neq i} \sum_{k, \ell} \pi_{ik} \pi_{j \ell} A_{ij}}{\sum_{j \neq i} \sum_{k, \ell} \pi_{ik} \pi_{j \ell} \hat{\theta}_{k \ell} \hat{\omega}_j}$.\;
    }
  }
\end{algorithm}

\begin{algorithm}
  \label{algo:pabm-em}
  \setstretch{1.5}
  \DontPrintSemicolon
  \SetAlgoLined
  \KwData{Adjacency matrix $A$, number of communities $K$}
  \KwResult{Estimated community label probabilities $\{\pi_{ik}\}$ for which each $\pi_{ik} = P(z_i = k \mid A)$, estimated popularity parameters $\{\hat{\lambda}_{ik}\}_K$.}
  \caption{Approximate EM algorithm for the PABM.}
  Initialize $\{\pi_{ik}\}$, $\{\hat{\lambda}_{ik}\}_K$.\;
  \While{$\|\nabla \ell\| > \epsilon$} {
    \For {$i = 1, ..., n$} {
      \For {$k = 1, ..., K$} {
      E-step: $\pi_{ik} \propto \exp \bigg( \sum_{j \neq i} \sum_{\ell} \pi_{j \ell} (A_{ij} \log (\hat{\lambda}_{ik} \hat{\lambda}_{j \ell}) + (1 - A_{ij}) \log (1 - \hat{\lambda}_{ik} \hat{\lambda}_{j \ell})) \bigg)$.\;
      M-step: $\hat{\lambda}_{ik} = \frac{\sum_{j \neq i} \sum_\ell \pi_{ik} \pi_{j \ell} A_{ij}}{\sum_{j \neq i} \sum_\ell \pi_{ik} \pi_{j \ell} \hat{\lambda}_{j \ell}}$.\;
      }
    }
  }
\end{algorithm}

@@

\begin{example}
\label{ex:uk-faculty-dcbm-pabm}
Recall the UK faculty friendship graph from example~\ref{ex:uk-faculty}. 
In example~\ref{ex:uk-faculty-sbm}, we fit an SBM to this graph using an EM algorithm (algorithm~\ref{algo:sbm-em}). 
\end{example}

@@

\citet{doi:10.1080/01621459.2012.699795} and \citet{rubindelanchy2017statistical} showed that the ASE is consistent in estimating the latent vectors, up to an indefinite orthogonal transformation. 
More specifically, for sparsity parameters such that $n \rho_n = \omega (\log^{4 c} n)$ for some constant $c > 1$ and indefinite orthogonal transformation $Q \in \mathbb{O}(p, q)$, 
$$\max_i \|Q \hat{x}_i - x_i \| = O_P \bigg( \frac{\log^c n}{n^{1/2}} \bigg)$$
where $x^\top$ and $\hat{x}^\top$ are the rows of $\hat{X}$ (from the ASE) and $X$, respectively.

@@

\begin{example}
Recall the UK faculty network in example \ref{ex:uk-faculty}. 
In examples~\ref{ex:uk-faculty-sbm} and \ref{ex:uk-faculty-dcbm-pabm}, we fit an SBM, DCBM, and PABM to the graph using their corresponding EM algorithms. 
While all three algorithms were able to correctly identify the labels for most of the vertices, it is not clear which model is the most appropriate for this graph. 
Analyzing the graph as a GRDPG, however, provides a visual indication for which block model is the best fit. 
If we wish to analyze this as an SBM or DCBM under the GRDPG framework, we can take the ASE of the adjacency matrix and see whether the embedding vectors lie around point masses or rays. 
We assume that this model is assortative, so $p = K$ and $q = 0$ in constructing the ASE. 
Because the fourth community only contains two members, we omitted it from this analysis. 
Figure \ref{fig:ukfaculty-ase} suggests that the appropriate model here is the DCBM, since each community appears to be described by a line segment. 
$K$-means clustering on the normalized embedding vectors results in a community detection accuracy of 96.7\%, which is an improvement over all three block model EM algorithms.

@@

\section*{Appendix A: Proofs of Theorems from Section 3}

@@

\section*{Appendix B: Proofs of Theorems from Section 4}

@@

\section*{Appendix C: Details on Fitting Bezier Curves to Noisy Data}

@@

\section*{Appendix D: Supplementary Materials}

>>>>

```{r}
import::from(magrittr, `%>%`, `%<>%`)
import::from(foreach, foreach, `%dopar%`)
library(ggplot2)
library(ggraph)
library(mclust)
source('~/dev/pabm-grdpg/functions.R')
source('~/dev/manifold-block-models/functions.R')
source('~/dev/dissertation/em-algorithms.R')

@@

\begin{algorithm}
  \label{algo:sbm-em}
  \setstretch{1.5}
  \DontPrintSemicolon
  \SetAlgoLined
  \KwData{Adjacency matrix $A$, number of communities $K$}
  \KwResult{Estimated community label probabilities $\{\pi_{ik}\}$ for which each $\pi_{ik} = P(z_i = k \mid A)$, estimated community edge probabilities $\{\hat{\theta}_{k \ell}\}_K$}
  \caption{Approximate EM algorithm for the SBM}
  Initialize $\{\pi_{ik}\}$, $\{\theta_{k \ell}\}$.\;
  \While{$\|\nabla \ell\| > \epsilon$} {
    \For {$i = 1, ..., n$} {
      \For {$k = 1, ..., K$} {
      E-step: $\pi_{ik} \propto \exp \bigg( \sum_{j \neq i} \sum_{\ell} \pi_{j \ell} (A_{ij} \log \hat{\theta}_{k \ell} + (1 - A_{ij}) \log (1 - \hat{\theta}_{k \ell})) \bigg)$.\;
      }
    }
    \For {$k = 1, ..., K$} {
      \For {$\ell = 1, ..., K$} {
        M-step: $\hat{\theta}_{k \ell} = \frac{\sum_{i < j} A_{ij} \pi_{ik} \pi_{j \ell}}{\sum_{i < j} \pi_{ik} \pi_{j \ell}}$.\;
      }
    }
  }
\end{algorithm}

@@

\begin{example}
\label{ex:uk-faculty-sbm}
Recall the UK faculty friendship network from example~\ref{ex:uk-faculty}. 
Suppose we assume that this graph is an SBM. 
Then applying algorithm \ref{algo:sbm-em} to this network yields a community detection accuracy of 79.7\%. 
For this analysis, the school with two faculty members is omitted (so $K=3$ and $n = 79$), and the EM algorithm was initialized with random $\{\pi_{ik}\}$, $\hat{\theta}_{k k} = 0.9$, and $\hat{\theta}_{k \ell} = 0.1$ for $k \neq \ell$. 
```{r cache = TRUE}
data(UKfaculty, package = 'igraphdata')
A <- as.matrix(igraph::as_adjacency_matrix(UKfaculty, type = 'both'))
A <- sign(A + t(A))
z <- factor(igraph::vertex_attr(UKfaculty)$Group)
A <- A[z != 4, z != 4]
z <- z[z != 4]
K <- 3
n <- length(z)

set.seed(123)
out <- em.sbm(A, 
              t(apply(matrix(runif(n * K), nrow = n), 
                      1, function(x) x / sum(x))), 
              matrix(c(.9, .1, .1, .1, .9, .1, .1, .1, .9), nrow = 3))
```
\end{example}

@@

\begin{algorithm}
  \label{algo:dcbm-em}
  \setstretch{1.5}
  \DontPrintSemicolon
  \SetAlgoLined
  \KwData{Adjacency matrix $A$, number of communities $K$}
  \KwResult{Estimated community label probabilities $\{\pi_{ik}\}$ for which each $\pi_{ik} = P(z_i = k \mid A)$, estimated community edge probabilities $\{\hat{\theta}_{k \ell}\}_K$, estimated degree factors $\hat{\omega}$.}
  \caption{Approximate EM algorithm for the DCBM.}
  Initialize $\{\pi_{ik}\}$, $\{\hat{\theta}_{k \ell}\}_K$, $\hat{\omega}$.\;
  \While{$\|\nabla \ell\| > \epsilon$} {
    \For {$i = 1, ..., n$} {
      \For {$k = 1, ..., K$} {
      E-step: $\pi_{ik} \propto \exp \bigg( \sum_{j \neq i} \sum_{\ell} \pi_{j \ell} (A_{ij} \log (\hat{\theta}_{k \ell} \hat{\omega}_i \hat{\omega}_j) + (1 - A_{ij}) \log (1 - \hat{\theta}_{k \ell} \hat{\omega}_i \hat{\omega}_j)) \bigg)$.\;
      }
    }
    \For {$k = 1, ..., K$} {
      \For {$\ell = 1, ..., K$} {
        M-step for $\theta_{k \ell}$: $\hat{\theta}_{k \ell} = \frac{\sum_{i < j} A_{ij} \pi_{ik} \pi_{j \ell}}{\sum_{i < j} \pi_{ik} \pi_{j \ell} \hat{\omega}_i \hat{\omega}_j}$.\;
      }
    }
    \For {$i = 1, ..., n$} {
      M-step for $\omega_i$: $\hat{\omega}_i = \frac{\sum_{j \neq i} \sum_{k, \ell} \pi_{ik} \pi_{j \ell} A_{ij}}{\sum_{j \neq i} \sum_{k, \ell} \pi_{ik} \pi_{j \ell} \hat{\theta}_{k \ell} \hat{\omega}_j}$.\;
    }
  }
\end{algorithm}

\begin{algorithm}
  \label{algo:pabm-em}
  \setstretch{1.5}
  \DontPrintSemicolon
  \SetAlgoLined
  \KwData{Adjacency matrix $A$, number of communities $K$}
  \KwResult{Estimated community label probabilities $\{\pi_{ik}\}$ for which each $\pi_{ik} = P(z_i = k \mid A)$, estimated popularity parameters $\{\hat{\lambda}_{ik}\}_K$.}
  \caption{Approximate EM algorithm for the PABM.}
  Initialize $\{\pi_{ik}\}$, $\{\hat{\lambda}_{ik}\}_K$.\;
  \While{$\|\nabla \ell\| > \epsilon$} {
    \For {$i = 1, ..., n$} {
      \For {$k = 1, ..., K$} {
        E-step: $\pi_{ik} \propto \exp \bigg( \sum_{j \neq i} \sum_{\ell} \pi_{j \ell} (A_{ij} \log (\hat{\lambda}_{ik} \hat{\lambda}_{j \ell}) + (1 - A_{ij}) \log (1 - \hat{\lambda}_{ik} \hat{\lambda}_{j \ell})) \bigg)$.\;
      }
    }
    \For {$i = 1, ..., n$} {
      \For {$k = 1, ..., K$} {
        M-step: $\hat{\lambda}_{ik} = \frac{\sum_{j \neq i} \sum_\ell \pi_{ik} \pi_{j \ell} A_{ij}}{\sum_{j \neq i} \sum_\ell \pi_{ik} \pi_{j \ell} \hat{\lambda}_{j \ell}}$.\;
      }
    }
  }
\end{algorithm}

@@

\begin{example}
\label{ex:uk-faculty-dcbm-pabm}
Recall the UK faculty friendship graph from example~\ref{ex:uk-faculty}. 
In example~\ref{ex:uk-faculty-sbm}, we fit an SBM to this graph using an EM algorithm (algorithm~\ref{algo:sbm-em}), yielding a community detection accuracy of 79.7\%. 
Starting with the output of algorithm~\ref{algo:sbm-em} and applying algorithm ~\ref{algo:dcbm-em} yields a greater accuracy of 94.9\%. 
This is expected, since the SBM is a special case of the DCBM, i.e., the SBM is the DCBM with some of the parameters restricted. 
Therefore, it is not immediately obvious from the improved accuracy whether the DCBM is a better fit for this graph than the SBM. 
Looking at the dispersion of the fitted degree factors $\hat{\omega}$, there does appear to be some indication that the DCBM is a better fit (figure \ref{fig:uk-faculty-dcbm-omega}). 
```{r uk-faculty-dcbm-omega, cache = TRUE, fig.width = 8, fig.height = 2, fig.cap = 'Histogram of the fitted degree factors for the UK faculty friendship network assuming that it is a DCBM. The DCBM was fit to this graph using algorithm \\ref{algo:dcbm-em}. The dispersion of the degree factors suggests that the DCBM is a better fit than the SBM.'}
data(UKfaculty, package = 'igraphdata')
A <- as.matrix(igraph::as_adjacency_matrix(UKfaculty, type = 'both'))
A <- sign(A + t(A))
z <- factor(igraph::vertex_attr(UKfaculty)$Group)
A <- A[z != 4, z != 4]
z <- z[z != 4]
K <- 3
n <- length(z)

set.seed(123)
out <- em.sbm(A, 
              t(apply(matrix(runif(n * K), nrow = n), 1, function(x) x / sum(x))),
              matrix(c(.9, .1, .1, .1, .9, .1, .1, .1, .9), nrow = K))
out <- em.dcbm(A, 
               out$pi,
               out$theta,
               runif(n))

ggplot() + 
  geom_histogram(aes(x = out$omega),
                 colour = 'black', fill = 'white') + 
  labs(x = 'Fitted degree factors for the UK faculty network') + 
  theme_bw()
```
\end{example}

@@

\citet{doi:10.1080/01621459.2012.699795} and \citet{rubindelanchy2017statistical} showed that the ASE is consistent in estimating the latent vectors, up to an indefinite orthogonal transformation. 
More specifically, for sparsity parameters such that $n \rho_n = \omega (\log^{4 c} n)$ for some constant $c > 1$ and indefinite orthogonal transformation $Q \in \mathbb{O}(p, q)$, 
$$\max_i \|Q \hat{x}_i - x_i \| = O_P \bigg( \frac{\log^c n}{n^{1/2}} \bigg)$$
where $x_i^\top$ and $\hat{x}_i^\top$ are the rows of $\hat{X}$ (from the ASE) and $X$, respectively.

@@

\begin{example}
Recall the UK faculty network in example \ref{ex:uk-faculty}. 
In examples~\ref{ex:uk-faculty-sbm} and \ref{ex:uk-faculty-dcbm-pabm}, we fit an SBM, DCBM, and PABM to the graph using their corresponding EM algorithms. 
While all three algorithms were able to correctly identify the labels for most of the vertices, it is not immediately clear which model is the most appropriate for this graph. 
Analyzing the graph as a GRDPG, however, provides a visual indication for which block model is the best fit. 
If we wish to analyze this as an SBM or DCBM under the GRDPG framework, we can take the ASE of the adjacency matrix and see whether the embedding vectors lie around point masses or rays. 
We assume that this model is assortative, so $p = K$ and $q = 0$ in constructing the ASE. 
Because the fourth community only contains two members, we omitted it from this analysis. 
Figure \ref{fig:ukfaculty-ase} suggests that the appropriate model here is the DCBM, since each community appears to be described by a line segment. 
$K$-means clustering on the normalized embedding vectors results in a community detection accuracy of 96.7\%, which is an improvement over all three block model EM algorithms.

@@

\section{Appendix A: Proofs of Theorems from Section 3}

@@

\section{Appendix B: Proofs of Theorems from Section 4}

@@

\section{Appendix C: Details on Fitting Bezier Curves to Noisy Data}

@@

\section{Appendix D: Supplementary Materials}

### 2022-11-12 draft/draft.tex
still writing

\usepackage{amsthm}
\newtheorem{theorem}{Theorem}[section]
\newtheorem{lemma}{Lemma}[section]
\newtheorem{corollary}{Corollary}[section]
\newtheorem{proposition}{Proposition}[section]
\newtheorem{conjecture}{Conjecture}[section]
\theoremstyle{definition}
\newtheorem{definition}{Definition}[section]
\theoremstyle{definition}
\newtheorem{example}{Example}[section]
\theoremstyle{definition}
\newtheorem{exercise}{Exercise}[section]
\theoremstyle{definition}
\newtheorem{hypothesis}{Hypothesis}[section]
\theoremstyle{remark}
\newtheorem*{remark}{Remark}
\newtheorem*{solution}{Solution}
\begin{document}

@@

\begin{example}
\label{ex:uk-faculty-sbm}
Recall the UK faculty friendship network from example~\ref{ex:uk-faculty}. 
\end{example}

@@

\begin{example}
\label{ex:uk-faculty-dcbm-pabm}
Recall the UK faculty friendship graph from example~\ref{ex:uk-faculty}. 
In example~\ref{ex:uk-faculty-sbm}, we fit an SBM to this graph using an EM algorithm (algorithm~\ref{algo:sbm-em}). 
\end{example}

@@

\citet{doi:10.1080/01621459.2012.699795} and \citet{rubindelanchy2017statistical} showed that the ASE is consistent in estimating the latent vectors, up to an indefinite orthogonal transformation.
More specifically, for sparsity parameters such that \(n \rho_n = \omega (\log^{4 c} n)\) for some constant \(c > 1\) and indefinite orthogonal transformation \(Q \in \mathbb{O}(p, q)\),
\[\max_i \|Q \hat{x}_i - x_i \| = O_P \bigg( \frac{\log^c n}{n^{1/2}} \bigg)\]
where \(x^\top\) and \(\hat{x}^\top\) are the rows of \(\hat{X}\) (from the ASE) and \(X\), respectively.

>>>>

\usepackage{amsthm}
\newtheorem{theorem}{Theorem}[section]
\newtheorem{lemma}{Lemma}[section]
\newtheorem{corollary}{Corollary}[section]
\newtheorem{proposition}{Proposition}[section]
\newtheorem{conjecture}{Conjecture}[section]
\theoremstyle{definition}
\newtheorem{definition}{Definition}[section]
\theoremstyle{definition}
\newtheorem{example}{Example}[section]
\theoremstyle{definition}
\newtheorem{exercise}{Exercise}[section]
\theoremstyle{definition}
\newtheorem{hypothesis}{Hypothesis}[section]
\theoremstyle{remark}
\newtheorem{remark}{Remark}
\newtheorem*{solution}{Solution}
\begin{document}

@@

\begin{example}
\label{ex:uk-faculty-sbm}
Recall the UK faculty friendship network from example~\ref{ex:uk-faculty}. 
Suppose we assume that this graph is an SBM. 
Then applying algorithm \ref{algo:sbm-em} to this network yields a community detection accuracy of 79.7\%. 
For this analysis, the school with two faculty members is omitted (so $K=3$ and $n = 79$), and the EM algorithm was initialized with random $\{\pi_{ik}\}$, $\hat{\theta}_{k k} = 0.9$, and $\hat{\theta}_{k \ell} = 0.1$ for $k \neq \ell$.

@@

\begin{example}
\label{ex:uk-faculty-dcbm-pabm}
Recall the UK faculty friendship graph from example~\ref{ex:uk-faculty}. 
In example~\ref{ex:uk-faculty-sbm}, we fit an SBM to this graph using an EM algorithm (algorithm~\ref{algo:sbm-em}), yielding a community detection accuracy of 79.7\%. 
Starting with the output of algorithm~\ref{algo:sbm-em} and applying algorithm ~\ref{algo:dcbm-em} yields a greater accuracy of 94.9\%. 
This is expected, since the SBM is a special case of the DCBM, i.e., the SBM is the DCBM with some of the parameters restricted. 
Therefore, it is not immediately obvious from the improved accuracy whether the DCBM is a better fit for this graph than the SBM. 
Looking at the dispersion of the fitted degree factors $\hat{\omega}$, there does appear to be some indication that the DCBM is a better fit (figure \ref{fig:uk-faculty-dcbm-omega}). 
\begin{figure}[H]

{\centering \includegraphics{draft_files/figure-latex/uk-faculty-dcbm-omega-1}

\caption{Histogram of the fitted degree factors for the UK faculty friendship network assuming that it is a DCBM. The DCBM was fit to this graph using algorithm \ref{algo:dcbm-em}. The dispersion of the degree factors suggests that the DCBM is a better fit than the SBM.}\label{fig:uk-faculty-dcbm-omega}
\end{figure}
\end{example}

@@

\citet{doi:10.1080/01621459.2012.699795} and \citet{rubindelanchy2017statistical} showed that the ASE is consistent in estimating the latent vectors, up to an indefinite orthogonal transformation.
More specifically, for sparsity parameters such that \(n \rho_n = \omega (\log^{4 c} n)\) for some constant \(c > 1\) and indefinite orthogonal transformation \(Q \in \mathbb{O}(p, q)\),
\[\max_i \|Q \hat{x}_i - x_i \| = O_P \bigg( \frac{\log^c n}{n^{1/2}} \bigg)\]
where \(x_i^\top\) and \(\hat{x}_i^\top\) are the rows of \(\hat{X}\) (from the ASE) and \(X\), respectively.
